Solution to ``Chapter 2: Basic tail and concentration bounds"

### 2020-07-25 skills/math/tail_bound/tailbound_solution.tex
tail bound

@@

\begin{defn}[Bernstein's condition]
Let $X$ be a random variable with mean $\mu = \bbE[X]$ and variance $ \sigma^2 = \var(X)$. We say $X$ satisfies the Bernstein's condition with parameter $b$ if
\begin{equation}
	\left| \bbE [(X -\mu)^k ]\right| \leq \frac{1}{2} k! \sigma^2 b^{k-2}, \quad \text{for } k =3, 4,....
\end{equation} 
\end{defn}
Note that bounded random variables satisfy the Bernstein's condition.

\begin{thm}[Bernstein-type bound] For any variable $X$ satisfying the Bernstein's condition, we have
\begin{equation}
	\bbE\left[e^{\lambda (X-\mu)}\right] \leq \exp\left\{ \frac{\lambda^2 \sigma^2}{2(1 - b |\lambda|)} \right\},\quad \text{for all } |\lambda| \leq \frac{1}{b}, 
\end{equation}
and the concentration inequality
\begin{equation}
	\bbP\left[|X - \mu| \geq t\right] \leq 2 \exp \left\{ -\frac{t^2}{2(\sigma^2 + bt)} \right\}, \quad \text{for all }t \geq 0.
\end{equation}

@@

>>>>

\newcommand{\of}[1]{\left(#1\right)}
\newcommand{\off}[1]{\left[#1\right]}
\newcommand{\offf}[1]{\left\{#1\right\}}
\newcommand{\aabs}[1]{\left|#1\right|}

@@

\begin{defn}[Bernstein's condition]
Let $X$ be a random variable with mean $\mu = \bbE[X]$ and variance $ \sigma^2 = \var(X)$. We say $X$ satisfies the Bernstein's condition with parameter $b$ if
\begin{equation}\label{eq:berncond}
	\left| \bbE [(X -\mu)^k ]\right| \leq \frac{1}{2} k! \sigma^2 b^{k-2}, \quad \text{for } k =3, 4,....
\end{equation} 
\end{defn}
Note that bounded random variables satisfy the Bernstein's condition.

\begin{thm}[Bernstein-type bound] For any variable $X$ satisfying the Bernstein's condition, we have
\begin{equation}
	\bbE\left[e^{\lambda (X-\mu)}\right] \leq \exp\left\{ \frac{\lambda^2 \sigma^2}{2(1 - b |\lambda|)} \right\},\quad \text{for all } |\lambda| \leq \frac{1}{b}, 
\end{equation}
and the concentration inequality
\begin{equation}\label{eq:thmbernconc}
	\bbP\left[|X - \mu| \geq t\right] \leq 2 \exp \left\{ -\frac{t^2}{2(\sigma^2 + bt)} \right\}, \quad \text{for all }t \geq 0.
\end{equation}

@@

\subsection{Exercise 2.8}
\begin{lem}[Bernstein and expectation] Let $Z$ be a nonnegative random variable satisfying the following concentration inequality
\begin{equation}\label{eq:berncond1}
	\bbP[Z \geq t] \leq C e^{- \frac{t^2}{2(\nu^2 +B t)}}, \quad \text{for all } t \geq 0,
\end{equation}
	where $(\nu, B)$ are two positive constants and $C \geq 1$. Then, the expectation of $Z$ satisfies 
	\begin{equation}\label{eq:bernexpec}
		\bbE[Z] \leq 2 \nu (\sqrt{\pi} + \sqrt{\log C}) + 4B (1 + \log C).
	\end{equation}
	Further, let $\{ X_i\}_{i=1}^n$ be a sequence of i.i.d.\ zero-mean variables satisfying the Bernstein condition \eqref{eq:berncond}. The sample mean of $\{ X_i\}_{i=1}^n$  satisfies
	\begin{equation}\label{eq:e28sample}
		\bbE \off{ \aabs{\frac{1}{n} \sum_{i=1}^n X_i }} \leq 2 \sigma (\sqrt{\pi} + \sqrt{\log 2}) + 4b (1 + \log 2).
	\end{equation}
\end{lem}

\begin{proof}
	First, we prove the equation~\eqref{eq:bernexpec}.

	\vspace{0.2cm}
	By equation~\eqref{eq:berncond1}, we have
	\begin{align}
		\bbP[Z \geq t] &\leq C \max \offf{\exp \of{-\frac{t^2}{4 \nu^2}}, \exp \of{-\frac{t^2}{4 Bt}} }\\
		&\leq C \exp \of{-\frac{t^2}{4 \nu^2}} + C \exp \of{-\frac{t^2}{4 Bt}}.\label{eq:e28tail}
	\end{align}

	Plugging the inequality~\eqref{eq:e28tail} to $\bbE[Z] = \int_{0}^{+\infty} \bbP[Z \geq t] dt$, we have
	\begin{equation}
		\bbE[Z] = \int_{0}^{+ \infty} \min \offf{1, C \exp \of{-\frac{t^2}{4 \nu^2}}} dt + \int_{0}^{+ \infty} \min\offf{1, C \exp \of{-\frac{t^2}{4 Bt}}} dt   \colonequals I_1 + I_2.
	\end{equation}

	To evaluate $I_1$, we spilt the integral to avoid the minimization. Solving $1 = C \exp \of{-\frac{t^2}{4 \nu^2}} $, the minimization term becomes
	\begin{align}
		\min \offf{1, C \exp \of{-\frac{t^2}{4 \nu^2}}} = \begin{cases}
			\quad 1 & \text{when } t \leq 2\nu \sqrt{\log C},\\
			\quad C \exp \of{-\frac{t^2}{4 \nu^2}} & \text{when } t \geq 2\nu \sqrt{\log C}.
		\end{cases}
	\end{align}
	Therefore, we evaluate $I_1$ as follows,
	\begin{align}
		I_1 &= \int_{0}^{ 2\nu \sqrt{\log C}} 1 dt + \int_{2\nu \sqrt{\log C}}^{+\infty} C \exp \of{-\frac{t^2}{4 \nu^2}} dt\\
		(\text{let}\quad y = \frac{t}{2\nu} - \sqrt{\log C})\quad &=  2\nu \sqrt{\log C} + 2\nu\int_{0}^{+\infty} \exp \of{-y^2 - 2y \sqrt{\log C}} dy\\
		&\leq  2\nu \sqrt{\log C} + 2\nu\int_{0}^{+\infty} \exp \of{-y^2} dy.
	\end{align}
	By Gaussian integral $\int_{0}^{+\infty} e^{-x^2} dx = \frac{\sqrt{\pi}}{2}$, we obtain $I_1 \leq 2\nu (\sqrt{\pi} + \sqrt{\log C})$. Similarly, we evaluate $I_2$ as follows 
	\begin{align}
		I_2 &= \int_{0}^{4B \log C} 1 dt + \int_{4B \log C}^{+\infty} C \exp \of{-\frac{t}{4B}} dt\\
		 &= 4B (\log C + 1).
	\end{align}
	Hence, we obtain the expectation of $Z$,
	\begin{equation}
		\bbE[Z] = I_1 + I_2 \leq  2\nu (\sqrt{\pi} + \sqrt{\log C}) + 4B (\log C + 1).
	\end{equation}

	\vspace{0.2cm}
	Next, we prove the equation~\eqref{eq:e28sample}.

	\vspace{0.2cm}
	For all $i \in [n]$, since $X_i$ satisfies the Bernstein condition with parameter $(\sigma,b)$, the variable $X_i$ satisfies the concentration bound~\eqref{eq:thmbernconc},
	\begin{equation}
		\bbP[|X_i| \geq t] \leq 2 \exp \offf{- \frac{t^2}{2(\sigma^2 + bt)}}, \quad \text{for all }t \geq 0. 
	\end{equation}
	By equation~\eqref{eq:bernexpec}, we have
	\begin{equation}
		\bbE[|X_i|] \leq  2 \sigma (\sqrt{\pi} + \sqrt{\log 2}) + 4b (1 + \log 2).
	\end{equation}

	Therefore, the expectation of the sample mean satisfies
	\begin{equation}
		\bbE \off{ \aabs{\frac{1}{n} \sum_{i=1}^n X_i }} \leq \frac{1}{n} \sum_{i=1}^n \bbE[|X_i|] \leq 2 \sigma (\sqrt{\pi} + \sqrt{\log 2}) + 4b (1 + \log 2).
	\end{equation}

\end{proof}

\begin{rmk}
	For all nonnegative random variables satisfying the Bernstein-type inequality \eqref{eq:berncond1} with parameters $(\nu,B,C)$, the expectations of the variables are upper bounded by a function of $(\nu,B,C)$. Particularly, for a zero-mean random variable $X$ satisfying the Bernstein condition with parameter $b$, $|X|$ satisfies the inequality \eqref{eq:berncond1} with parameters $(\sigma,b,2)$, where $\sigma^2 = \var(X)$.  The expectation of the absolute variable $|X|$ depends on the variance $\sigma^2$ and the parameter $b$.
\end{rmk}
